Gammapy: A Python package for gamma-ray astronomy

### 2023-07-25 src/ms-review.tex
Fix typos and remove comments

%% Context
%% CN: I think Fermi-LAT data were available from the 2nd year on
%% https://fermi.gsfc.nasa.gov/ssc/data/policy/
While \fermi data and analysis tools have been public from early years
of the project \citep{Atwood2009}, ground-based \gammaray astronomy has been historically conducted
through experiments operated by independent collaborations, each relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been so far successful, it does not
permit easy combination of data from several instruments. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially in light of the fact that different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of energy range covered,
duty cycle and spatial resolution.

%TODO: better transition into context
% CN: last-minute transition improvement
The Cherenkov Telescope Array Observatory (CTAO) will be the first ground-based
\gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by \cta will be reconstructed event lists and corresponding instrument response functions.}
will be shared publicly after some proprietary period, along with the software required to analyze them.
The adoption of an open and standardised \gammaray data format, beside being a necesity for future observatories such as \cta,
will be greatly beneficial to the current generation of instruments, eventually allowing to exploit their data legacies even
beyond the end of their scientific operations.

@@

%% Context: development of open software
%% CN: I don't see the parallel between gamma-ray astronomy establishing itself and the success of python
%% maybe this was the introduction of the very first draft?
%% I would just write: "In the past decade, \textit{Python} has become..."
In the past decade \textit{Python} has become extremely popular as a scientific programming language,
in particular in the field of data sciences. This success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a \enquote{glue} language between different programming languages and last but not least
the rich ecosystem of packages and its open and supportive community \citep{Momcheva2015}.
In the sub-field of computational astronomy, the \astropy project \citep{astropy} was created in 2012
to build a community-developed core Python package for astronomy.
It offers basic functionalities that astronomers of many fields need, such as representing
and transforming astronomical coordinates, manipulating physical quantities including units
as well as reading and writing FITS files.

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project was started following the model of \astropy, with the objective of building a common
software library for \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. The various \gammaray instruments
can export their data to a common format (\gadf) and then combine and analyse
them using a common software library.
The \gammapy package is an independent community-developed software project.
It has been selected to be the core library for the Science Analysis tools of \cta,
but also involves contributors associated to other instruments.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

%% CN: what's this??? Out of place here! %%%%
%They consist of two
%observations of $20\,{\rm min}$ each, chosen from the dataset used to estimate
%the performance of the upgraded stereo system~\citep{magic_performance} and
%already included in~\cite{joint_crab}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

With the public availability of the \gadf format specifications and the
\gammapy package, some experiments started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf format \citep{HESS_DR1} for data level 3 (DL3) \gammaray data.
This data release served as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.][]{Mohrmann2019}. Two observations of the Crab nebula have
been released by the \magic collaboration~\citep{magic_performance}.
Using these public data from \fermi, \hess, \magic and additional observations
provided by FACT and \veritas, the authors of \cite{joint_crab} presented
a combined analysis of \gammaray data from different instruments for the first time.
Later the \hawc collaboration also released a limited test dataset of the Crab Nebula,
which was used to validate the \gammapy package in \cite{Olivera2022}.
%% CN: I would remark this also demonstrated Gammapy could be applied to WCD
%% depite having been designed for IACTs initially
The increased availability of public data that followed the definition of
a common data format, and the development of \gammapy as a community-driven open software,
led the way toward a more open science in the very-high-energy \gammaray astronomy domain.
The adoption of \gammapy as science tools strengthens the commitment of the future \cta Observatory to the FAIR principles
(Findable, Accessible, Interoperable and Reusable) principles \citep{FAIR16, FAIR22} 
that define the key requirements for open science.

@@

\section{Gamma-ray Data Analysis}
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and names given as the corresponding Python code suffix, e.g. 
		\code{gammapy.data}. Below each icon there is a list of the most
        important objects defined in the sub-package. The light grey folder
		icons show the subpackages for the most fundamental data structures such 
		as maps and IRFs. The bottom of the figure shows the high-level analysis
		sub-module with its dependcy on the YAML file format. 
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
The data analysis process in \gammaray astronomy is usually split into two stages.
The first one deals with the data processing from detector measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

@@

%% CN: ref. for King function https://ui.adsabs.harvard.edu/abs/2011A%26A...534A..34R/abstract
%% CN: I would not say that the PSFKing takes into account "the gamma and sigma", fo what?
%% I would put a formula or quickly describe the function (e.g. it is a power-law, an exponential...)
\subsubsection{Point Spread Function}
\gammapy allows users to treat different kinds of PSFs,
in particular, parametric multi-dimensional Gaussian distributions (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing}). The \code{EnergyDependentMultiGaussPSF}
class is able to handle up to three Gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. The general \code{ParametricPSF} class allows users to create a
custom PSF with a parametric representation different from Gaussian(s) or King profile(s).
The generic \code{PSF3D} class stores a radial symmetric profile of a
PSF to represent non-parametric shapes, again depending on true energy
and offset from the pointing position.

@@

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow a later access by name in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automtically resolved
        by \gammapy. The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_datasets}.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with the associated models and likelihood function, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modelling and fitting purposes. Depending upon the type of analysis and the
associated statistic, different types of Datasets are supported. The \code{MapDataset} is
used for combined spectral and morphological (3D) fitting, while a 1D spectral 
fitting can be performed using the \code{SpectrumDataset}.
While the default fit statistics for both of these classes is the \emph{Cash}~\citep{Cash}
statistic, there are other classes which support
analyses where the background is measured from control regions, so called \enquote{off} obervations.
Those require the use of a different fit statistics, which takes into account the
uncertainty of the background measurement. This case is covered by the \code{MapDatasetOnOff}
and \code{SpectrumDatasetOnOff} classes, which use the \emph{WStat}~\citep{WStat} statistic.

@@

The correction of background models from the data themselves is supported 
by specific \code{Maker} classes such as the \code{FoVBackgroundMaker} or the
\code{ReflectedRegionsBackgroundMaker}. The former is used to estimate the
normalisation of the background model from the data themselves, while the
latter is used to estimate teh background from regions reflected from the
poiting position.
%% CN: I would add a single line explaining what each of them do

@@

\begin{figure}
	\small
	\import{code-examples/generated/}{gp_stats}
	\caption{
        Using \code{gammapy.stats} to compute statistical quantities
        such as excess, significance and asymetric errors
        from counts based data. The data array such as \code{counts}, \code{counts\_off}
		and the background efficency ratio \code{alpha} are passed on initialisation
        of the \code{WStatCountsStatistic} class. The derived quantities
        are then computed dynamically from the corresponding class
        attributes such as \code{stat.n\_sig} for the excess
		and \code{stat.sqrt\_ts} for the significance.
		The output of the code example is shown in Figure~\ref{fig:code_example_gp_stats}.
    }
	\label{fig*:minted:gp_stats}
\end{figure}

@@

The models follow a naming scheme which contains the category as a suffix to
the class name. The spectral models include a special class of normed models,
named using the \code{NormSpectralModel} suffix.
These spectral models feature a dimension-less \textit{norm} parameter
instead of an \textit{amplitude} parameter with physical units. They
can be used as an energy-dependent multiplicative correction
factor to another spectral model. They are typically used for
adjusting template-based models, or, for example, to take into account
the absorbtion effect on \gammaray spectra caused by the extra-galactic background
light (EBL) (\code{EBLAbsorptionNormSpectralModel}). \gammapy supports a variety
of EBL absorption models, such as those from \cite{Franceschini2008, Franceschini2017}, \cite{Finke2010},
and \cite{Dominguez2011}.

@@

%% CN: should we mention that the dataset also include the stat function?
%% and thus the fit interfaces only with the dataset
The \code{Fit} class provides methods to fit, i.e. optimise, model parameters and estimate
their errors and correlations. It interfaces with a \code{Datasets} object, which
in turn is connected to a \code{Models} object containing the model parameters in its
\code{Parameters} object. Models can be unique for a given dataset, or contribute to
multiple datasets, allowing e.g., to perform a joint fit to
multiple IACT datasets, or to jointly fit IACT and \fermi datasets. Many
examples are given in the tutorials.
%% CN: I would write "to perform a joint fit from multiple datasets from the same instruments
%% (e.g. collected with different configurations) or from multiple datasets from different gamma-ray
%% instruments (space- and ground-based)."

@@

Note that, for now, covariance matrix and errors are computed only for the fitting with 
\iminuit. However, depending on
the problem other optimizers can perform better, so sometimes it can be useful
to run a pre-fit with alternative optimization methods. In the future, we plan to
extend the supported fitting backends, including for example solutions based on Markov chain Monte Carlo methods.
\footnote{a prototype is available in gammapy-recipes,
	\url{https://gammapy.github.io/gammapy-recipes/_build/html/notebooks/mcmc-sampling-emcee/mcmc_sampling.html}
}
%% CN: watchout with giving links to notebooks whose position might change or disappear in the future.

@@

Both result objects support the possibility to serialise
the data into multiple formats. This includes the
\gadf SED format \footnote{\url{https://gamma-astro-data-formats.readthedocs.io/en/latest/spectra/flux_points/index.html}},
FITS-based ND sky maps and other formats compatible with Astropy's \code{Table} and
\code{BinnedTimeSeries} data structures. This allows
users to further analyse the results with Astropy, for example using
standard algorithms for time analysis, such as
the Lomb-Scargle periodogram or the Bayesian
blocks. So far, \gammapy does not support unfolding of \gammaray spectra.
Methods for this will be implemented in future versions of \gammapy.
%% CN: reference for unfolding https://ui.adsabs.harvard.edu/abs/2007NIMPA.583..494A/abstract

@@

The \code{Analysis} class has the responsibility for orchestrating the workflow
defined in the configuration \code{AnalysisConfig} objects and triggering the execution of
the \code{AnalysisStep} classes that define the identified common use cases. These
steps include the following: observations selection with the \code{DataStore},  data
reduction, excess map computation, model fitting, flux points estimation, and
light curves production.
%% CN: I would not say that these are use cases, rather analysis or data reduction steps

@@

\section{Applications}
\label{sec:applications}
%% CN: I think we never explained what gammapy-data is and we should have at this point 
%% (it is used without mention also in the TSEstimator example).
\gammapy is currently used for a variety of analyses by different IACT
experiments and has already been employed in about 65 scientific publications as of 21/03/2023
\footnote{\href{https://ui.adsabs.harvard.edu/search/q=(\%20(citations(doi\%3A\%2210.1051\%2F0004-6361\%2F201834938\%22)\%20OR\%20citations(bibcode\%3A2017ICRC...35..766D))\%20AND\%20year\%3A2014-2023)&sort=date\%20desc\%2C\%20bibcode\%20desc&p_=0}{List on ADS}}.
In this section, we illustrate the capabilities of \gammapy by performing some standard
analysis cases commonly considered in \gammaray astronomy.
Beside reproducing standard methodologies, we illustrate the unique data combination
capabilites of \gammapy by presenting a multi-instrument analysis, which is not possible within any
of the current instrument private software frameworks.
The examples shown are based on the data accessible in the \code{gammapy-data} repository,
and limited by the availability of public data.
We remark that, as long as the data are compliant with the GADF specifications (or its future evolutions),
and hence with Gammapy's data structures, there is no limitation on performing
analyses of data from a given instrument.

\subsection{1D Analysis}
\label{ssec:1d-analysis}
One of the most common analysis cases in \gammaray astronomy is measuring the
spectrum of a source in a given region defined on the sky, in conventional
astronomy also called \textit{aperture photometry}. The spectrum is typically measured
in two steps: first a parametric spectral model is fitted to the data and
secondly flux points are computed in a pre-defined set of energy bins. The
result of such an analysis performed on three simulated CTA observations is
shown in Figure~\ref{fig:cta_galactic_center}. In this case the spectrum was
measured in a circular aperture centered on the Galactic Center, in
\gammaray~astronomy often called \enquote{on region}. For such analysis the user first
chooses a region of interest and energy binning, both defined by a
\code{RegionGeom}. In a second step, the events and the IRFs are binned
into maps of this geometry, by the \code{SpectrumDatasetMaker}. All the data and
reduced IRFs are bundled into a \code{SpectrumDataset}. To estimate
the expected background in the \enquote{on region} a \enquote{reflected regions} background
method was used~\citep{Berge07}, represented in \gammapy by the
\code{ReflectedRegionsBackgroundMaker} class. The resulting reflected regions are
illustrated for all three observations overlayed on the counts map in Figure~\ref{fig:cta_galactic_center}.
After reduction, the data were modelled using a forward-folding method and assuming
a point source with a power law spectral shape. The model was defined, using
the \code{SkyModel} class with a \code{PowerLawSpectralModel} spectral component only.
This model was then combined with the \code{SpectrumDataset}, which contains the reduced data 
and fitted using the \code{Fit} class. Based on this best-fit model, the final flux points and corresponding
log-likelihood profiles were computed using the \code{FluxPointsEstimator}. The
example takes $<10$ seconds to run on a standard laptop with M1 arm64 CPU.

\begin{figure*}
	\centering
	\includegraphics[width=1.\textwidth]{figures/cta_galactic_center.pdf}
	\caption{
		Example of a one dimensional spectral analysis of the Galactic Center for three simulated
		observations from the first CTA data challenge. The left image shows the maps of counts with the signal
		region in white and the reflected background regions for the three different observations
		overlaid in different colors. The right image shows the resulting spectral flux points and
		their corresponding log-likelihood profiles. The flux points are shown in orange, with the
		horizontal bar illustrating the width of the energy bin and the vertical bar the $1~\sigma$
		error. The log-likelihood profiles for each enetgy bin are shown in the background. The colormap
		illustrates the difference of the log-likelihood to the log-likelihood of the best fit value.
		}
	\label{fig:cta_galactic_center}
\end{figure*}
%% CN: expand 1DC acronym and add a reference, here or in the section text

@@

Combining the datasets in a \code{Datasets} list, \gammapy automatically generates
a likelihood including three different types of terms, two Poissonian likelihoods
for \fermi's \code{MapDataset} and MAGIC's \code{SpectrumDataset}, and a $\chi^2$
accounting for the \hawc flux points. For \fermi, a three-dimensional forward folding
of the sky model with the IRF is performed, in order to compute the predicted counts
in each sky-coordinate and energy bin. For \magic, a one-dimensional forward-folding
of the spectral model with the \irfs is performed to predict the counts in each estimated energy bin. A log
parabola is fitted over almost five decades in energy \SI{10}{GeV} to \SI{300}{TeV}, taking into account all flux points from all three datasets.
%% CN: is the LP formula specified elsewhere?

@@

In short, the production of catalogs from \gammaray surveys can be divided in
four main steps: (a) data reduction, (b) object detection, (c) model fitting and model
selection and (d) associations and classification. All steps can either be done directly
with \gammapy or by relying on the seamless integration of \gammapy with the
scientific Python ecosystem. This allows to rely on 3rd party functionality
wherever needed. A simplifed catalog analysis based on \gammapy typically includes:

@@

\gammapy has been succesfully used for catalog studies performed on simulations
of the future \cta Galactic Plane Survey~\citep{Remy2021}. Besides the scientific
insights they also gave us the opportunity to test the \gammapy software on complex
use cases. This resulted in a number of improvements to the \gammapy package, such as
as improved performance, optimized analysis strategies documented to users, and
identifying the needs for future developments, including solutions for distributed
computing. The catalog studies of the \cta Galactic Plane Survey simulations
also allowed for detailed cross-comparison of the results obtained with the independent
\gammaray analysis package \textit{ctools}\citep{2016A&A...593A...1K}, with very consistent 
results.

@@

\gammapy is an international open-source project with a broad
developer base and contributions and commitments from mutiple groups and
leading institutes in the very high-energy astrophysics
domain\footnote{\url{https://gammapy.org/team.html}}. The main development
roadmaps are discussed and validated by a \textit{Coordination Committee}, composed of
representatives of the main contributing institutions and observatories.
This committee is
chaired by a \textit{Project Manager} and his deputy while two \textit{Lead Developers} manage
the development strategy and organise technical activities. This
institutionally-driven organisation, the permanent staff and commitment of
supporting institutes ensure the continuity of the executive teams. A core team
of developers from the contributing institutions is in charge of the regular
development, which benefits from regular contributions of the community at
large.

>>>>

%% Context
While \fermi data and analysis tools have been public from early years
of the project \citep{Atwood2009}, ground-based \gammaray astronomy has been historically conducted
through experiments operated by independent collaborations, each relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been so far successful, it does not
permit easy combination of data from several instruments. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially in light of the fact that different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of energy range covered,
duty cycle and spatial resolution.

%TODO: better transition into context
The Cherenkov Telescope Array Observatory (CTAO) will be the first ground-based
\gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by \cta will be reconstructed event lists and corresponding instrument response functions.}
will be shared publicly after some proprietary period, along with the software required to analyze them.
The adoption of an open and standardised \gammaray data format, beside being a necesity for future observatories such as \cta,
will be greatly beneficial to the current generation of instruments, eventually allowing to exploit their data legacies even
beyond the end of their scientific operations.

@@

%% Context: development of open software
In the past decade \textit{Python} has become extremely popular as a scientific programming language,
in particular in the field of data sciences. This success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a \enquote{glue} language between different programming languages and last but not least
the rich ecosystem of packages and its open and supportive community \citep{Momcheva2015}.
In the sub-field of computational astronomy, the \astropy project \citep{astropy} was created in 2012
to build a community-developed core Python package for astronomy.
It offers basic functionalities that astronomers of many fields need, such as representing
and transforming astronomical coordinates, manipulating physical quantities including units
as well as reading and writing FITS files.

%% \gammapy: concept and goals
The \gammapy project was started following the model of \astropy, with the objective of building a common
software library for \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. The various \gammaray instruments
can export their data to a common format (\gadf) and then combine and analyse
them using a common software library.
The \gammapy package is an independent community-developed software project.
It has been selected to be the core library for the Science Analysis tools of \cta,
but also involves contributors associated to other instruments.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

With the public availability of the \gadf format specifications and the
\gammapy package, some experiments started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf format \citep{HESS_DR1} for data level 3 (DL3) \gammaray data.
This data release served as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.][]{Mohrmann2019}. Two observations of the Crab nebula have
been released by the \magic collaboration~\citep{magic_performance}.
Using these public data from \fermi, \hess, \magic and additional observations
provided by FACT and \veritas, the authors of \cite{joint_crab} presented
a combined analysis of \gammaray data from different instruments for the first time.
Later the \hawc collaboration also released a limited test dataset of the Crab Nebula,
which was used to validate the \gammapy package in \cite{Olivera2022}.

The increased availability of public data that followed the definition of
a common data format, and the development of \gammapy as a community-driven open software,
led the way toward a more open science in the very-high-energy \gammaray astronomy domain.
The adoption of \gammapy as science tools strengthens the commitment of the future \cta Observatory to the FAIR principles
(Findable, Accessible, Interoperable and Reusable) principles \citep{FAIR16, FAIR22} 
that define the key requirements for open science.

@@

\section{Gamma-ray Data Analysis}
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and names given as the corresponding Python code suffix, e.g. 
		\code{gammapy.data}. Below each icon there is a list of the most
        important objects defined in the sub-package. The light grey folder
		icons show the subpackages for the most fundamental data structures such 
		as maps and IRFs. The bottom of the figure shows the high-level analysis
		sub-module with its dependency on the YAML file format. 
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
The data analysis process in \gammaray astronomy is usually split into two stages.
The first one deals with the data processing from detector measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

@@

\subsubsection{Point Spread Function}
\gammapy allows users to treat different kinds of PSFs,
in particular, parametric multi-dimensional Gaussian distributions (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing}). The \code{EnergyDependentMultiGaussPSF}
class is able to handle up to three Gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. The general \code{ParametricPSF} class allows users to create a
custom PSF with a parametric representation different from Gaussian(s) or King profile(s).
The generic \code{PSF3D} class stores a radial symmetric profile of a
PSF to represent non-parametric shapes, again depending on true energy
and offset from the pointing position.

@@

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow a later access by name in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automatically resolved
        by \gammapy. The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_datasets}.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with the associated models and likelihood function, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modelling and fitting purposes. Depending upon the type of analysis and the
associated statistic, different types of Datasets are supported. The \code{MapDataset} is
used for combined spectral and morphological (3D) fitting, while a 1D spectral 
fitting can be performed using the \code{SpectrumDataset}.
While the default fit statistics for both of these classes is the \emph{Cash}~\citep{Cash}
statistic, there are other classes which support
analyses where the background is measured from control regions, so called \enquote{off} observations.
Those require the use of a different fit statistics, which takes into account the
uncertainty of the background measurement. This case is covered by the \code{MapDatasetOnOff}
and \code{SpectrumDatasetOnOff} classes, which use the \emph{WStat}~\citep{WStat} statistic.

@@

The correction of background models from the data themselves is supported 
by specific \code{Maker} classes such as the \code{FoVBackgroundMaker} or the
\code{ReflectedRegionsBackgroundMaker}. The former is used to estimate the
normalisation of the background model from the data themselves, while the
latter is used to estimate the background from regions reflected from the
pointing position.

@@

\begin{figure}
	\small
	\import{code-examples/generated/}{gp_stats}
	\caption{
        Using \code{gammapy.stats} to compute statistical quantities
        such as excess, significance and asymmetric errors
        from counts based data. The data array such as \code{counts}, \code{counts\_off}
		and the background efficiency ratio \code{alpha} are passed on initialisation
        of the \code{WStatCountsStatistic} class. The derived quantities
        are then computed dynamically from the corresponding class
        attributes such as \code{stat.n\_sig} for the excess
		and \code{stat.sqrt\_ts} for the significance.
		The output of the code example is shown in Figure~\ref{fig:code_example_gp_stats}.
    }
	\label{fig*:minted:gp_stats}
\end{figure}

@@

The models follow a naming scheme which contains the category as a suffix to
the class name. The spectral models include a special class of normed models,
named using the \code{NormSpectralModel} suffix.
These spectral models feature a dimension-less \textit{norm} parameter
instead of an \textit{amplitude} parameter with physical units. They
can be used as an energy-dependent multiplicative correction
factor to another spectral model. They are typically used for
adjusting template-based models, or, for example, to take into account
the absorption effect on \gammaray spectra caused by the extra-galactic background
light (EBL) (\code{EBLAbsorptionNormSpectralModel}). \gammapy supports a variety
of EBL absorption models, such as those from \cite{Franceschini2008, Franceschini2017}, \cite{Finke2010},
and \cite{Dominguez2011}.

@@

The \code{Fit} class provides methods to fit, i.e. optimise, model parameters and estimate
their errors and correlations. It interfaces with a \code{Datasets} object, which
in turn is connected to a \code{Models} object containing the model parameters in its
\code{Parameters} object. Models can be unique for a given dataset, or contribute to
multiple datasets, allowing e.g., to perform a joint fit to
multiple IACT datasets, or to jointly fit IACT and \fermi datasets. Many
examples are given in the tutorials.

@@

Note that, for now, covariance matrix and errors are computed only for the fitting with 
\iminuit. However, depending on
the problem other optimizers can perform better, so sometimes it can be useful
to run a pre-fit with alternative optimization methods. In the future, we plan to
extend the supported fitting backends, including for example solutions based on Markov chain Monte Carlo methods.
\footnote{a prototype is available in gammapy-recipes,
	\url{https://gammapy.github.io/gammapy-recipes/_build/html/notebooks/mcmc-sampling-emcee/mcmc_sampling.html}
}

@@

Both result objects support the possibility to serialise
the data into multiple formats. This includes the
\gadf SED format \footnote{\url{https://gamma-astro-data-formats.readthedocs.io/en/latest/spectra/flux_points/index.html}},
FITS-based nd sky maps and other formats compatible with Astropy's \code{Table} and
\code{BinnedTimeSeries} data structures. This allows
users to further analyse the results with Astropy, for example using
standard algorithms for time analysis, such as
the Lomb-Scargle periodogram or the Bayesian
blocks. So far, \gammapy does not support unfolding of \gammaray spectra.
Methods for this will be implemented in future versions of \gammapy.

@@

The \code{Analysis} class has the responsibility for orchestrating the workflow
defined in the configuration \code{AnalysisConfig} objects and triggering the execution of
the \code{AnalysisStep} classes that define the identified common use cases. These
steps include the following: observations selection with the \code{DataStore},  data
reduction, excess map computation, model fitting, flux points estimation, and
light curves production.

@@

\section{Applications}
\label{sec:applications}
\gammapy is currently used for a variety of analyses by different IACT
experiments and has already been employed in about 65 scientific publications as of 21/03/2023
\footnote{\href{https://ui.adsabs.harvard.edu/search/q=(\%20(citations(doi\%3A\%2210.1051\%2F0004-6361\%2F201834938\%22)\%20OR\%20citations(bibcode\%3A2017ICRC...35..766D))\%20AND\%20year\%3A2014-2023)&sort=date\%20desc\%2C\%20bibcode\%20desc&p_=0}{List on ADS}}.
In this section, we illustrate the capabilities of \gammapy by performing some standard
analysis cases commonly considered in \gammaray astronomy.
Beside reproducing standard methodologies, we illustrate the unique data combination
capabilities of \gammapy by presenting a multi-instrument analysis, which is not possible within any
of the current instrument private software frameworks.
The examples shown are based on the data accessible in the \code{gammapy-data} repository,
and limited by the availability of public data.
We remark that, as long as the data are compliant with the GADF specifications (or its future evolutions),
and hence with Gammapy's data structures, there is no limitation on performing
analyses of data from a given instrument.

\subsection{1D Analysis}
\label{ssec:1d-analysis}
One of the most common analysis cases in \gammaray astronomy is measuring the
spectrum of a source in a given region defined on the sky, in conventional
astronomy also called \textit{aperture photometry}. The spectrum is typically measured
in two steps: first a parametric spectral model is fitted to the data and
secondly flux points are computed in a pre-defined set of energy bins. The
result of such an analysis performed on three simulated CTA observations is
shown in Figure~\ref{fig:cta_galactic_center}. In this case the spectrum was
measured in a circular aperture centered on the Galactic Center, in
\gammaray~astronomy often called \enquote{on region}. For such analysis the user first
chooses a region of interest and energy binning, both defined by a
\code{RegionGeom}. In a second step, the events and the IRFs are binned
into maps of this geometry, by the \code{SpectrumDatasetMaker}. All the data and
reduced IRFs are bundled into a \code{SpectrumDataset}. To estimate
the expected background in the \enquote{on region} a \enquote{reflected regions} background
method was used~\citep{Berge07}, represented in \gammapy by the
\code{ReflectedRegionsBackgroundMaker} class. The resulting reflected regions are
illustrated for all three observations overlaid on the counts map in Figure~\ref{fig:cta_galactic_center}.
After reduction, the data were modelled using a forward-folding method and assuming
a point source with a power law spectral shape. The model was defined, using
the \code{SkyModel} class with a \code{PowerLawSpectralModel} spectral component only.
This model was then combined with the \code{SpectrumDataset}, which contains the reduced data 
and fitted using the \code{Fit} class. Based on this best-fit model, the final flux points and corresponding
log-likelihood profiles were computed using the \code{FluxPointsEstimator}. The
example takes $<10$ seconds to run on a standard laptop with M1 arm64 CPU.

\begin{figure*}
	\centering
	\includegraphics[width=1.\textwidth]{figures/cta_galactic_center.pdf}
	\caption{
		Example of a one dimensional spectral analysis of the Galactic Center for three simulated
		observations from the first CTA data challenge. The left image shows the maps of counts with the signal
		region in white and the reflected background regions for the three different observations
		overlaid in different colors. The right image shows the resulting spectral flux points and
		their corresponding log-likelihood profiles. The flux points are shown in orange, with the
		horizontal bar illustrating the width of the energy bin and the vertical bar the $1~\sigma$
		error. The log-likelihood profiles for each enetgy bin are shown in the background. The colormap
		illustrates the difference of the log-likelihood to the log-likelihood of the best fit value.
		}
	\label{fig:cta_galactic_center}
\end{figure*}

@@

Combining the datasets in a \code{Datasets} list, \gammapy automatically generates
a likelihood including three different types of terms, two Poissonian likelihoods
for \fermi's \code{MapDataset} and MAGIC's \code{SpectrumDataset}, and a $\chi^2$
accounting for the \hawc flux points. For \fermi, a three-dimensional forward folding
of the sky model with the IRF is performed, in order to compute the predicted counts
in each sky-coordinate and energy bin. For \magic, a one-dimensional forward-folding
of the spectral model with the \irfs is performed to predict the counts in each estimated energy bin. A log
parabola is fitted over almost five decades in energy \SI{10}{GeV} to \SI{300}{TeV}, taking into account all flux points from all three datasets.

@@

In short, the production of catalogs from \gammaray surveys can be divided in
four main steps: (a) data reduction, (b) object detection, (c) model fitting and model
selection and (d) associations and classification. All steps can either be done directly
with \gammapy or by relying on the seamless integration of \gammapy with the
scientific Python ecosystem. This allows to rely on 3rd party functionality
wherever needed. A simplified catalog analysis based on \gammapy typically includes:

@@

\gammapy has been successfully used for catalog studies performed on simulations
of the future \cta Galactic Plane Survey~\citep{Remy2021}. Besides the scientific
insights they also gave us the opportunity to test the \gammapy software on complex
use cases. This resulted in a number of improvements to the \gammapy package, such as
as improved performance, optimized analysis strategies documented to users, and
identifying the needs for future developments, including solutions for distributed
computing. The catalog studies of the \cta Galactic Plane Survey simulations
also allowed for detailed cross-comparison of the results obtained with the independent
\gammaray analysis package \textit{ctools}\citep{2016A&A...593A...1K}, with very consistent 
results.

@@

\gammapy is an international open-source project with a broad
developer base and contributions and commitments from multiple groups and
leading institutes in the very high-energy astrophysics
domain\footnote{\url{https://gammapy.org/team.html}}. The main development
roadmaps are discussed and validated by a \textit{Coordination Committee}, composed of
representatives of the main contributing institutions and observatories.
This committee is
chaired by a \textit{Project Manager} and his deputy while two \textit{Lead Developers} manage
the development strategy and organise technical activities. This
institutionally-driven organisation, the permanent staff and commitment of
supporting institutes ensure the continuity of the executive teams. A core team
of developers from the contributing institutions is in charge of the regular
development, which benefits from regular contributions of the community at
large.
